Static Race Detection in OS Kernels by Mining Locking Rules

### 2023-07-26 3.technique.tex
Update technique

\PP{S3: Locking-rule deduction.} After collecting all accesses to key fields, 
the analysis deduces locking rules based on statistics. However, on the one 
hand, distinguishing different accesses to the same data structure field by 
different program sites is not fine-grained enough, because the access to a 
variable and the acquirement to a specific lock are often packaged in a special 
function, and all accesses under different calling context are regarded as the 
same access. On the other hand, distinguishing different accesses by different 
code paths can also suffer from inaccuracy when a function contains many branch 
statement and cause numerous code paths for the same variable access. Based on 
these consideration, the analysis distinguishes different accesses to the same 
data structure field by different calling contexts. Specifically, given a key 
field {\em f} and a lock {\em l}, the analysis first finds all access to {\em 
f} with lock {\em l} from the collected field accesses in Stage 2, and gets the 
number {\em locked\_access\_num} by counting different calling contexts 
extracted from the code paths of these field accesses. Then, the analysis finds 
all access to {\em f} (no matter which lock is accessed), and gets the number 
{\em all\_access\_num} in the same way as {\em locked\_access\_num}. If the 
ratio {\em locked\_access\_num}/{\em all\_access\_num} is larger than a given 
threshold, and these is at least one write access to {\em f}, the data 
structure field {\em f} is inferred to be protected by the lock {\em l}.

@@

After extracting key fields, the method collects all accesses to key fields 
with lock set analysis and the alias graph. Take the code path {\tt Line 15, 
Line 6, Line 7, Line 8, Line 9, Line 16 and Line 17} as an example, through a 
lock instruction at Line 6, the analysis records the acquired lock {\em 
Dev.lock} into the lock set. When analyzing the read instruction at Line 7, the 
analysis finds the node of the accessed field ($\mathit{n_2}$) and the node of 
the lock stored in the lock set ($\mathit{n_4}$) has the same ancestor 
($\mathit{n_1}$), and thus records a field access <{\em Dev.flag}, {\em 
Dev.lock}>. Similar to the read instruction at Line 7, the method also records 
a field access <{\em Dev.flag*}, {\em Dev.lock}> for the write instruction at 
Line 8. Then through an unlock operation at Line 9, the analysis removes the 
lock {\em Dev.lock} is removed from the lock set. As a result, the key fields 
{\em Dev.flag} and {\em Dev.flag*} are accessed without acquiring any lock, and 
thus the analysis records two field accesses <{\em Dev.flag}, {\em NULL}> and 
<{\em Dev.flag*}, {\em NULL}>. The final field accesses are shown in 
Figure~\ref{fig_demo_rule_mining}.

After collecting all field accesses, the analysis finds that the fields {\em 
Dev.flag} and {\em Dev.flag*} are accessed under three calling contexts, and 
two of them are protected by the lock {\em Dev.lock}. And thus the two fields 
{\em Dev.flag} and {\em Dev.flag*} is deduce to be protected by the lock {\em 
Dev.lock} (assume the threshold of the ratio locked\_access\_num / 
all\_access\_num is 0.6).

\subsection{Lock-Usage Analysis}
\label{subsec_lock_usage_analysis}

>>>>

\PP{S3: Locking-rule deduction.} After collecting all accesses to key fields, 
the analysis deduces locking rules based on statistics. However, on the one 
hand, distinguishing different accesses to the same data structure field by 
different program sites is not fine-grained enough, because the access to a 
variable and the acquirement to a specific lock are often packaged in a special 
function, and all accesses under different calling context are regarded as the 
same access. On the other hand, distinguishing different accesses by different 
code paths can also suffer from inaccuracy when a function contains many branch 
statement and cause numerous code paths for the same variable access. Based on 
these consideration, the analysis distinguishes different accesses to the same 
data structure field by different calling contexts. Specifically, given a key 
field {\em f} and a lock {\em l}, the analysis first finds all access to {\em 
f} with lock {\em l} from the collected field accesses in Stage 2, and gets the 
number {\em locked\_access\_num} by counting different calling contexts 
extracted from the code paths of these field accesses. Then, the analysis finds 
all access to {\em f} (no matter which lock is accessed), and gets the number 
{\em all\_access\_num} in the same way as {\em locked\_access\_num}. If the 
ratio {\em locked\_access\_num}/{\em all\_access\_num} is larger than a given 
threshold, and these is at least one write access to {\em f}, the data 
structure field {\em f} is inferred to be protected by the lock {\em l}. After 
deducting locking rules, the analysis detecting data races by checking whether 
a field access validates any locking rules.

@@

After extracting key fields, the method collects all accesses to key fields 
with lock set analysis and the alias graph. Take the code path {\tt Line 15, 
Line 6, Line 7, Line 8, Line 9, Line 16 and Line 17} as an example, through a 
lock instruction at Line 6, the analysis records the acquired lock {\em 
Dev.lock} into the lock set. When analyzing the read instruction at Line 7, the 
analysis finds the node of the accessed field ($\mathit{n_2}$) and the node of 
the lock stored in the lock set ($\mathit{n_4}$) has the same ancestor 
($\mathit{n_1}$), and thus records a field access <{\em Dev.flag}, {\em 
Dev.lock}>. Similar to the read instruction at Line 7, the method also records 
a field access <{\em Dev.flag*}, {\em Dev.lock}> for the write instruction at 
Line 8. Then through an unlock operation at Line 9, the analysis removes the 
lock {\em Dev.lock} from the lock set. As a result, the key fields {\em 
Dev.flag} and {\em Dev.flag*} are accessed without acquiring any lock, and thus 
the analysis records two field accesses <{\em Dev.flag}, {\em NULL}> and <{\em 
Dev.flag*}, {\em NULL}>. The final field accesses are shown in 
Figure~\ref{fig_demo_rule_mining}(c).

After collecting all field accesses, the analysis finds that the fields {\em 
Dev.flag} and {\em Dev.flag*} are accessed under three calling contexts, and 
two of them are protected by the lock {\em Dev.lock}. And thus the two fields 
{\em Dev.flag} and {\em Dev.flag*} is deduce to be protected by the lock {\em 
Dev.lock} (assume the threshold of the ratio {\em locked\_access\_num} / 
{\em all\_access\_num} is 0.6). However, the accesses at Lines 16 and 17 are 
not protected by {\em Dev.lock} and thus causing two data races.

\subsection{Lock-Usage Analysis}
\label{subsec_lock_usage_analysis}
Our analysis takes functions that has no caller function as the entry as 
existing work~\cite{Li:ASPLOS22}, and performs a path-based analysis. It 
assumes that all codes reaching from entry functions can execute concurrently 
to reduce false negatives. However, this assumption is too conservative and can 
introduce many false positives because not all entry functions can execute 
concurrently in fact. We observe that each kernel module has an initialization 
phase, after which many function can be called concurrently by other modules. 
When performing initialization, the kernel module serially initialized the lock 
and prepare other data for subsequent operations. And thus functions with the 
lock initialization and functions called by them tend not to be executed 
concurrently. Based on the observation, we propose a lock-usage analysis to 
filter out false positives caused by code that can not execute concurrently.

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/fig_pseudocode_lock_usage.pdf}
	\figcaption{Pseudocodes of lock-usage analysis.}
	\label{fig_pseudocode_lock_usage}
\end{figure}

Our lock-usage analysis uses the union-find set~\cite{Galler:ACM64} to get all 
functions that reachable from the lock initialization function such as {\em 
mutex\_init()} and {\em spin\_lock\_init()}. 
Figure~\ref{fig_pseudocode_lock_usage} shows the pseudocodes of the lock-usage 
analysis. For each analyzed function in the OS kernel, the analysis first gets 
all the called function of it (Lines 3-4), and then union the analyzed function 
and the called function to update the union-find set (Line 5). After processing 
all the call instructions, the analysis judging whether a function executes in 
the initialization phase by checking whether it exist in the same set with the 
given lock initialization function (Lines 9-11).

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/fig_demo_lock_usage.pdf}
	\figcaption{A false data race filtered out by our lock-usage analysis.}
	\label{fig_demo_lock_usage}
\end{figure}

\noindent{\textbf{\em Example.}} Figure~\ref{fig_demo_lock_usage} shows a false 
data race filtered out by our lock-usage analysis in the Linux DRM scheduler. 
For example in {\em drm\_sched\_entity\_set\_priority()}, the accesses to {\em 
entity->priority} is protected by the lock {\em entity->rq\_lock} in most 
cases. Therefore, {\em entity-priority} is deduced to be protected by the lock 
{\em entity->rq\_lock}. But in {\em drm\_sched\_entity\_init()}, {\em 
entity->priority} is written without acquiring the lock {\em entity->rq\_lock} 
and thus introducing a possible data race. However, the lock initialization 
function {\em spin\_lock\_init()} is called at Line 86 by {\em 
drm\_sched\_entity\_init()}. And thus {\em drm\_sched\_entity\_init()} is 
inferred to execute in the module initialization phase and can not execute 
concurrently with other function, and this is a false data race.

\subsection{Pattern-Based Estimation}
\label{subsec_estimation}
